---
title: "Lab 06 — Hypothesis Testing from Claims"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab introduces **hypothesis testing** as an extension of sampling, bootstrap distributions, and uncertainty quantification.

The core idea is simple:

> A **claim** about the population becomes a **null hypothesis**.

You will use the **US-AMS simplified AmeriFlux dataset** again and test claims about population means using:

-   one sample,
-   a bootstrap estimate of uncertainty,
-   a z-statistic,
-   and a p-value.

This lab emphasizes **reasoning from claims to tests**, not just mechanics.

------------------------------------------------------------------------

# Learning goals

By the end of this lab, you should be able to:

-   Turn a verbal claim into a null hypothesis
-   Choose a variable and define the relevant population parameter
-   Draw a sample and estimate a mean
-   Use bootstrap resampling to estimate uncertainty in that mean
-   Compute a z-statistic and p-value
-   Interpret a hypothesis test in plain language
-   Connect confidence intervals and hypothesis tests

------------------------------------------------------------------------

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=FALSE}
# GOAL: Read the dataset and clean the names.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023)

glimpse(df)
```

## Derive time columns (if needed)

```{r derive_time, echo=TRUE, eval=FALSE}
# GOAL: Create date / month / day_night columns as in earlier labs.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE),
    day_night = if_else(daytime == 1, "Day", "Night")
  )
```

------------------------------------------------------------------------

# Part I — From claim to hypothesis test

## 1) Choose a variable

Choose **one numeric variable** from the dataset.

Possible examples: - `swc` - `ta` - `fc` - `gpp` - `le`

```{r choose_variable, echo=TRUE, eval=FALSE}
target_var <- "le"

# CHECK: does this variable exist?
target_var %in% names(df)
```

## 2) Describe the population parameter

Write the population parameter you care about:

> $\mu$ = the true 2023 population mean of `latent heat flux`

Fill in the blank in words:

> In this lab, my population parameter is the true annual mean of **`latent heat flux`** at the US-AMS site.

------------------------------------------------------------------------

## 3) Make a claim

Write a claim about the population mean for your variable.

Examples: - “The annual mean soil water content is 0.30.” - “The annual mean air temperature is 10 °C.” - “The annual mean latent heat flux is 40 W m$^{-2}$.”

Now turn your claim into hypotheses:

-   Null hypothesis: $H_0: \mu = \mu_0$
-   Alternative hypothesis: $H_A: \mu \neq \mu_0$

Fill these in:

> My claim is: `The annual mean latent heat flux is 100 W m$^{-2}$.`

> $H_0: \mu = \mu_0$

> $H_A: \mu \neq \mu_0$

------------------------------------------------------------------------

## 4) Draw one sample and compute the sample mean

```{r draw_sample, echo=TRUE, eval=FALSE}
# GOAL: Draw one random sample and compute its mean.

set.seed(100)

n_samp <- 200   # e.g., 200

samp <- df %>%
  filter(!is.na(.data[[target_var]])) %>%
  slice_sample(n = n_samp)

sample_mean <- mean(samp[[target_var]], na.rm = TRUE)

sample_mean

```

**Prompt (2–3 sentences):** How does your sample mean compare to your claim? Is it close, or already quite different?

> *It's kind of close, considering the range of values we're working with. Latent heat flux has a range of ~550 units, so being only ~40 units off isn't too bad. It's not terribly different.*

------------------------------------------------------------------------

## 5) Bootstrap the sample mean

```{r bootstrap_means, echo=TRUE, eval=FALSE}
# GOAL: Bootstrap the sample mean to estimate uncertainty.

set.seed(100)

B <- 4000   # e.g., 4000

boot_means <- replicate(
  B,
  mean(sample(samp[[target_var]], size = n_samp, replace = TRUE), na.rm = TRUE)
)

head(boot_means)
```

## 6) Estimate uncertainty

```{r bootstrap_summary, echo=TRUE, eval=FALSE}
# GOAL: Compute the bootstrap SE and 95% CI.

boot_se <- sd(boot_means)
boot_ci <- quantile(boot_means, probs = c(0.025, 0.975))

tibble(
  sample_mean = sample_mean,
  boot_se = boot_se,
  ci_lower = boot_ci[1],
  ci_upper = boot_ci[2]
)
```

**Prompt (2–3 sentences):** What does the bootstrap SE tell you? What does the 95% CI tell you?

> *The bootstrap SE tells us the standard deviation of the 4000 bootstrap samples. The 95% CI tells us that 95% of the time, it is expected the 'true' value falls between 52.384 and 76.611. *

------------------------------------------------------------------------

## 7) Convert the claim into a formal test

Define the null value from your claim.

```{r hypothesis_test, echo=TRUE, eval=FALSE}
# GOAL: Compute z and p-value for your claim.

mu0 <- 100   # numeric claim value

z_stat <- (sample_mean - mu0) / boot_se

p_value <- 2 * pnorm(-abs(z_stat))

tibble(
  sample_mean = sample_mean,
  mu0 = mu0,
  boot_se = boot_se,
  z_stat = z_stat,
  p_value = p_value
)
```

------------------------------------------------------------------------

## 8) Interpret the result

Use your p-value and CI to answer:

-   Does the 95% CI include the null value?
-   Is the p-value small or large?
-   Would you reject the null hypothesis at $\alpha = 0.05$?

Fill in:

> The null value `100` is not inside my 95% CI. :(

> My p-value is `7.309943e-09`, which is small.

> Therefore, I would reject $H_0$ at the 0.05 level.

**Prompt (4–6 sentences):** Explain your decision in plain language. Do the data make the claim look plausible or surprising? Be careful: “fail to reject” does not mean “prove true.”

> *I said that the average latent heat flux is 100 Watts / square meter. However, the population mean is ~63 units and the 95% confidence interval is 52.38 - 76.61. This gives me a z-score of -5.78, which tells me 2 things: 1) The magnitude is large, meaning that my guess (100) is quite `off`, and 2) my guess is an overshoot (not undershoot) since the z score is negative.*
> *These data make my claim look surprising. Therefore, I reject the null hypothesis - the data do not support it.*

------------------------------------------------------------------------

## 9) Visualize the hypothesis test

```{r plot_test, echo=TRUE, eval=FALSE}
# GOAL: Plot the bootstrap distribution, sample mean, CI, and null value.

ggplot(tibble(boot_mean = boot_means), aes(x = boot_mean)) +
  geom_histogram(bins = 100, alpha = 0.8) +
  geom_vline(xintercept = sample_mean, linetype = "dashed", linewidth = 1.1) +
  geom_vline(xintercept = boot_ci, color = "red", linewidth = 1.1) +
  geom_vline(xintercept = mu0, color = "blue", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", target_var),
    y = "Count",
    title = "Bootstrap distribution and null claim",
    subtitle = "Dashed = sample mean; Red = 95% CI; Blue = null value"
  )
```

**Prompt (2–3 sentences):** Does the plot agree with your p-value-based conclusion? Explain.

> *Yes. It shows my guessed mean is outside the 95% CI and above the true mean (which is why the z-score was negative).*

------------------------------------------------------------------------

## 10) Try a second claim (short extension)

Keep the same variable, but test a **different null value**.

```{r second_claim, echo=TRUE, eval=FALSE}
# GOAL: Test a second claim using the same sample and bootstrap SE.

mu0_alt <- 50

z_stat_alt <- (sample_mean - mu0_alt) / boot_se
p_value_alt <- 2 * pnorm(-abs(z_stat_alt))

tibble(
  mu0_alt = mu0_alt,
  z_stat_alt = z_stat_alt,
  p_value_alt = p_value_alt
)
```

**Prompt (3–4 sentences):** How did changing the null value change the conclusion? What does this tell you about the relationship between the null hypothesis and the observed estimate?

> *I made the null value smaller (and closer to the population mean). This changed the sign of the z-score, decreased the z-score magnitude, and increased the p-value. We can no longer reject the null hypothesis; 50 could be the true pop. mean.*
>*This tells me that the null hypothesis value (50) is closer to the observed estimate (63).*

------------------------------------------------------------------------

# Part II — To be added later

We will add additional exercises here on:

-   comparing means between groups,
-   one- vs two-sided tests,
-   and connections to t-tests / ANOVA.

> *Instructor will add this section later. Do not delete this header.*

------------------------------------------------------------------------

# Submission

-   Knit your `.Rmd` to HTML.
-   Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
-   Submit the GitHub repository link on Canvas.
